Forcing for Isabelle

### 2018-12-30 LICS/intro_m.tex
frase sobre testear la tecnologia de prueba actual... No muy convencido

In this work we address a substantial part of formalizing the proof
that given a model $M$ of $\ZF$, any \emph{generic extension} $M[G]$
given by forcing also satisfies $\ZF$. As remarked by
\citet[][p.250]{kunen2011set} \enquote{[...] in verifying that $M[G]$
  is a model for set theory, the hardest axiom to verify is
  [Separation].}

\medskip
\fbox{(sigue p\'arrafo) Versi\'on de Pedro}
\medskip

The most important achievement of this paper is the
mechanization in the proof assistant \emph{Isabelle} of a proof of the
Axiom of Separation in 
generic extensions by using the ``fundamental'' theorems of forcing.
En route
to this, we also formalized the satisfaction by $M[G]$ of
Extensionality, Foundation, Union, and Infinity. % We have already
% proved Pairing in the first report of our project
% \cite{2018arXiv180705174G}.
We have  followed \cite{kunen2011set} for this purpose.

\medskip
\fbox{Versi\'on de Miguel}
\medskip

The most important achievement of this paper is the
mechanization in Isabelle of a considerable part of this result; en route
to this, we also formalized the satisfaction by $M[G]$ of
Extensionality, Foundation, Union, and Infinity. % We have already
% proved Pairing in the first report of our project
% \cite{2018arXiv180705174G}.

\medskip
\fbox{Fin p\'arrafo}
\medskip

Our development benefited from the remarkable work done by Lawrence
Paulson \cite{paulson_2003} on the formalization of G\"odel's
constructible universe in Isabelle. The
ultimate goal of our project is the formalization of the forcing
techniques needed to show the independence of the Continuum
Hypothesis. We think that this project constitutes an interesting case
which stresses how feasible is to formally implement mathematics that
involve several levels of reasoning. \fbox{No entiendo la ultima oraci\'on}

>>>>

In this work we address a substantial part of formalizing the proof
that given a model $M$ of $\ZF$, any \emph{generic extension} $M[G]$
given by forcing also satisfies $\ZF$. As remarked by
\citet[][p.250]{kunen2011set} \enquote{[...] in verifying that $M[G]$
  is a model for set theory, the hardest axiom to verify is
  [Separation].}  
The most important achievement of this paper is the
mechanization in the proof assistant \emph{Isabelle} of a proof of the
Axiom of Separation in 
generic extensions by using the ``fundamental'' theorems of forcing.
En route
to this, we also formalized the satisfaction by $M[G]$ of
Extensionality, Foundation, Union, and Infinity. % We have already
% proved Pairing in the first report of our project
% \cite{2018arXiv180705174G}.
We have  followed \cite{kunen2011set} for this purpose.

Our development benefited from the remarkable work done by Lawrence
Paulson \cite{paulson_2003} on the formalization of G\"odel's
constructible universe in Isabelle. The
ultimate goal of our project is the formalization of the forcing
techniques needed to show the independence of the Continuum
Hypothesis. 
We think that this project constitutes an interesting test-case
for the current technology of formalization of mathematics, in
particular for the need of handling several layers of reasoning.
